# Advanced Graphics

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

## Chapter 6 - *Advanced graphics*
####I'll be the first to confess that my re-write of the chapter on advanced graphics may be redundant, but what the heck. 

```{r preliminaries}
library("Lahman")
library("tidyverse"); options(dplyr.width = Inf)

set.seed(8761825)
devtools::session_info()
```

### Section 6.2 - the lattice package. I think the power of the lattice package comes from the facets (multi-pane images). We can use the facet() function in ggplot2 to the same effect.
```{r Sec_6.2}
# The Verlander dataset
load("Data/balls_strikes_count.RData") #I've placed the relevent .RData file in the Data directory.
str(verlander)
dim(verlander)

sampleRows <- verlander %>% 
  sample_n(20, replace = FALSE) # Note that the sample function in R has replace = FALSE as the default as well. 
sampleRows
```

#### Figure_6.1
```{r Fig_6.1}
ggplot(verlander, aes(speed)) + 
  theme_bw() + 
  geom_histogram(bins = 30) + 
  xlab("Speed") + 
  ylab("Count")

ggplot(verlander, aes(speed)) + 
  theme_bw() + 
  geom_line(stat = "density") + 
  ylab("Density") + 
  xlab("Speed")
```

#### Figure 6.2 - **facets!!!!**
```{r Fig_6.2}
ggplot(verlander, aes(speed)) + 
  theme_bw() + 
  geom_line(stat = "density") + 
  ylab("Density") + 
  xlab("Speed") + 
  facet_wrap(~ pitch_type, nrow = 5)
```

#### Figure 6.3
```{r Fig_6.3}
ggplot(verlander, aes(speed, lty = pitch_type)) + 
  theme_bw() + 
  geom_line(stat = "density") + 
  ylab("Density") + 
  xlab("Speed")
```

#### Figure 6.4
```{r Fig_6.4}
F4verl <- verlander %>% 
  filter(pitch_type == "FF") %>% 
  mutate(gameDay = as.integer(format(gamedate, format = "%j")))
head(F4verl)
dim(F4verl)

# Using dplyr rather than aggregate
dailySpeed <- F4verl %>% 
  select(gameDay, season, speed) %>% 
  group_by(gameDay, season) %>% 
  summarize(speed = mean(speed))
dim(dailySpeed)
head(dailySpeed)

ggplot(dailySpeed, aes(y = speed, x = gameDay)) + 
  theme_bw() + 
  facet_wrap(~ season, nrow = 2) + 
  geom_point(size = 2) + 
  ylab("Pitch Speed (MPH)") + 
  xlab("Day of Year")
```

#### Figure 6.5
```{r Fig_6.5}
speedFC <- verlander %>% 
  filter(pitch_type %in% c("FF", "CH"))
head(speedFC)

avgspeedFC <- speedFC %>% 
  select(pitch_type, season, speed) %>% 
  group_by(season, pitch_type) %>% 
  summarize(speed = mean(speed))
avgspeedFC

ggplot(avgspeedFC, aes(y = season, x = speed)) + 
  theme_bw() + 
  geom_text(aes(label = pitch_type)) + 
  ylab("Season") + 
  xlab("Speed")
```

#### Figure 6.6
```{r Fig_6.6}
avgSpeed <- F4verl %>% 
  select(pitches, season, speed) %>% 
  group_by(pitches, season) %>% 
  summarize(speed = mean(speed))
head(avgSpeed)

ggplot(avgSpeed, aes(y = speed, x = pitches)) + 
  theme_bw() + 
  facet_wrap(~ season, nrow = 2) + 
  geom_point(size = 1.5) + 
  geom_hline(aes(yintercept = mean(speed))) 
```

#### Figure 6.7
```{r Fig_6.7}
NoHit <- verlander %>% 
  filter(gamedate == "2011-05-07")
dim(NoHit)
head(NoHit)

f6.7 <- ggplot(NoHit, aes(x = px, y = pz)) + 
  theme_bw() + 
  facet_wrap(~ batter_hand, nrow = 1) + 
  geom_point(aes(shape = pitch_type), size = 2.5)
f6.7
```

#### Figure 6.8
```{r Fig_6.8}
f6.8 <- ggplot(NoHit, aes(x = px, y = pz)) + 
  theme_bw() + 
  facet_wrap(~ batter_hand, ncol = 1) + 
  geom_point(aes(shape = pitch_type), size = 2.5) + 
  xlim(-2.2, 2.2) + ylim(0, 4)
f6.8
```

#### Figure 6.9
```{r Fig_6.9}
f6.9 <- ggplot(NoHit, aes(x = px, y = pz)) + 
  theme_bw() + 
  facet_wrap(~ batter_hand, nrow = 1) + 
  geom_point(aes(shape = pitch_type), size = 2.5) + 
  xlim(-2.2, 2.2) + 
  ylim(0, 5) + 
  ylab("Vertical Location\n(ft. from ground)") + 
  xlab("Horizontal Location\n(ft. from middle of plate)") +
  labs(shape = "Pitch Type")
f6.9
```

#### Figure 6.10
```{r Fig_6.10}
topKzone <- 3.5
botKzone <- 1.6
inKzone <- -0.95
outKzone <- 0.95

f6.10 <- ggplot(NoHit, aes(x = px, y = pz)) + 
  theme_bw() + 
  facet_wrap(~ batter_hand, nrow = 1) + 
  geom_point(aes(shape = pitch_type), size = 2.5) + 
  xlim(-2.2, 2.2) + 
  ylim(0, 5) + 
  ylab("Vertical Location\n(ft. from ground)") + 
  xlab("Horizontal Location\n(ft. from middle of plate)") +
  labs(shape = "Pitch Type") + 
  annotate("rect", xmin = -0.95, xmax = 0.95, ymin = 1.6, ymax = 3.5, fill = "dodgerblue", alpha = 0.3) # I think that shading in the strike zone is WAY prettier than a dashed line box. 
f6.10
```

#### Figure 6.11
```{r Fig_6.11}
head(cabrera)
dim(cabrera)

ggplot(cabrera, aes(x = hitx, y = hity)) + 
  geom_point()
```

#### Figure 6.12
```{Fig_6.12}
ggplot(cabrera, aes(x = hitx, y = hity)) + 
  geom_point(aes(color = hit_outcome)) + 
  coord_equal()
```

#### Figure 6.13
```{r Fig_6.13}
ggplot(cabrera, aes(x = hitx, y = hity)) + 
  geom_point(aes(color = hit_outcome)) + 
  coord_equal() + 
  facet_wrap(~ season)

bases <- data.frame(x = c(0, 90 / sqrt(2), 0, -90 / sqrt(2), 0), y = c(0, 90 / sqrt(2), 2 * 90 / sqrt(2), 90 / sqrt(2), 0))
```

#### Figure 6.14
```{r Fig_6.14}
ggplot(cabrera, aes(x = hitx, y = hity)) + 
  geom_point(aes(color = hit_outcome)) + 
  coord_equal() + 
  facet_wrap(~ season) + 
  geom_path(aes(x = x, y = y), data = bases) + 
  geom_segment(x = 0, xend = 300, y = 0, yend = 300) + 
  geom_segment(x = 0, xend = -300, y = 0, yend = 300)
```

#### Figure 6.15
```{r Fig_6.15}
cabreraStretch <- cabrera %>% 
  filter(gamedate > "2012-08-31")

ggplot(cabreraStretch, aes(x = hitx, y = hity)) + 
  geom_point(aes(shape = hit_outcome, color = pitch_type, size = speed)) + 
  coord_equal() + 
  geom_path(aes(x = x, y = y), data = bases) + 
  guides(col = guide_legend(ncol = 2))+ 
  geom_segment(x = 0, xend = 300, y = 0, yend = 300) + 
  geom_segment(x = 0, xend = -300, y = 0, yend = 300)
```

#### Figure 6.16 **Note that recent ggplot2 versions use geom_hline**
```{r Fig_6.16}
ggplot(F4verl, aes(x = pitches, y = speed)) + 
  facet_wrap(~ season) + 
  geom_hline(aes(yintercept = mean(speed)), lty = 3) + 
  geom_point(aes(x = pitches, y = speed), data = F4verl[sample(1:nrow(F4verl), 1000), ]) + 
  geom_smooth(col = "black") + 
  geom_vline(aes(xintercept = 100), col = "black", lty = 2)
```

#### Figure 6.17
```{r Fig_6.17}
kZone <- data.frame(x = c(inKzone, inKzone, outKzone, outKzone, inKzone), y = c(botKzone, topKzone, topKzone, botKzone, botKzone))
kZone

# In my experience, setting the facet prior to calling geom is good practice
ggplot(F4verl, aes(x = px, y = pz)) + 
  facet_wrap(~ batter_hand) + 
  geom_point() + 
  coord_equal() + 
  geom_path(aes(x, y), data = kZone, lwd = 2, col = "white")
```

#### Figure 6.18
```{r Fig_6.18}
ggplot(F4verl, aes(x = px, y = pz)) + 
  facet_wrap(~ batter_hand) + 
  stat_binhex() + 
  coord_equal() + 
  geom_path(aes(x, y), data = kZone, lwd = 2, col = "white", alpha = 0.3)
```

#### Figure 6.19
```{r Fig_6.19}
Comerica <- jpeg::readJPEG("Data/Comerica.jpg")

ggplot(cabrera, aes(x = hitx, y = hity)) + 
  coord_equal() + 
  theme(panel.background = element_rect(fill = "white")) + 
  annotation_raster(Comerica, -310, 305, -100, 480) + 
  stat_binhex(alpha = 0.9, binwidth = c(5, 5)) + 
  scale_fill_gradient(low = "grey70", high = "black") + 
  ylab("") +
  xlab("")
```

### Chapter 6 exercises
1. **Location of Pitches for Left- and Right-Handed Batters** - Use a density plot to display the horizontal location of Justin Verlander’s pitchers by opponent’s handedness. Choose the conditioning and grouping variables so that one can easily detect the differences in location by handedness. Add a legend (if necessary) and vertical reference lines indicating the borders of the strike zone.

```{r CH6.Q1}
ggplot(verlander, aes(x = px)) + 
  theme_bw() + 
  geom_line(stat = "density", aes(color = batter_hand), size = 2) + 
  xlab("Horizontal location") + 
  ylab("Density") + 
  geom_vline(xintercept = -0.95, linetype = 2, size = 1.5) +
  geom_vline(xintercept = 0.95, linetype = 2, size = 1.5)
```
2. **Comparing Pitch Locations for Two Pitchers** - The sanchez data frame contains 2008-2012 PITCHf/x data for pitcher Jonathan Sanchez. The structure of this data frame is similar to the verlander data frame described in the chapter. Use a graphical display to compare the ability of Sanchez and Verlander in maintaining their fast-ball speed through the game. (See Sections 6.2.7 and 6.3.8.) Use either the lattice or ggplot2 graphics package and display the data either as a multipanel plot or a superposed lines plot

Use the following function to plot multiple images on one pane. [Link to source of code](www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/)
```{r multiplot}
multiplot <- function(..., plotlist = NULL, file, cols = 1, layout = NULL){
  library(grid)
  plots <- c(list(...), plotlist)
  numPlots <- length(plots)
  if(is.null(layout)){
    layout <- matrix(seq(1, cols * ceiling(numPlots / cols)), ncol = cols, nrow = ceiling(numPlots / cols))
  }
  if (numPlots == 1){
    print(plots[[1]])
  }
  else{
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    for(i in 1:numPlots){
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row, layout.pos.col = matchidx$col))
    }
  }
}  
```
```{r Ch6.Q2, fig.width = 10, fig.height = 5}
head(sanchez)
dim(sanchez)

Sanchez_FF <- sanchez %>% filter(pitch_type == "FF") %>% select(speed, pitches)

S_FF <- ggplot(Sanchez_FF, aes(x = pitches, y = speed)) +
  theme_bw() + 
  geom_point(size = 1.75, alpha = 0.4) + 
  geom_smooth(method = "lm", col = "red") +
  ggtitle("Sanchez fastball") +
  xlab("Pitches") +
  ylab("Speed (mph)") +
  ylim(85, 105) + 
  xlim(0, 135)

# Earlier we created the F4verl object, which contains the data we need. 

V_FF <- ggplot(F4verl, aes(x = pitches, y = speed)) +
  theme_bw() + 
  geom_point(size = 1.75, alpha = 0.4) + 
  geom_smooth(method = "lm", col = "red") +
  ggtitle("Verlander fastball") +
  xlab("Pitches") +
  ylab("") +
  ylim(85, 105) + 
  xlim(0, 135)

multiplot(S_FF, V_FF, cols = 2)
# Cool. Verlander's fastbal speed increases the more he throws and Sanchez's velocity decreases.
```


3. **Graphical View of the Speeds of Justin Verlander’s Fastballs** 
+ The cut function is useful for recoding a continuous variable into intervals. Use this function to categorize the pitches variable in the verlander data frame in groups of ten pitches.
```{r Ch5.Q3a}
Verl_cut <- cut(verlander$pitches, breaks = 10)
head(Verl_cut)
table(Verl_cut)
```
+ Use the bwplot function from the lattice package to produce a boxplot of Verlander’s four-seam fastball speed (use the F4verl data frame) for each ten-pitches group. Compare the information conveyed by the resulting chart with that of Figure 6.6.
```{r Ch5.Q3b}
# I'll use ggplot2 for this. 
VC <- verlander %>% mutate(pitch_bin = cut(verlander$pitches, breaks = 10))
head(VC)
str(VC)

ggplot(VC, aes(y = speed, x = pitch_bin)) + 
  geom_boxplot(fill = "grey", outlier.size = 2, stat = "boxplot") +
  theme_bw() +
  xlab("Number of pitches") + 
  ylab("Speed (mph)") + 
  theme(axis.text.x = element_text(angle = 90))
# You can really see that he dialed up the spped in his final pitches
```
4. **Exploring Miguel Cabrera’s Slugging Power**
+ Create a data frame by selecting, from the cabrera data frame, the instances where the hit outcome variable assumes the value H (for base hit).
```{r Ch6.Q4a}
Q4a <- cabrera %>% filter(hit_outcome == "H")
head(Q4a)
dim(Q4a)
```
+ Using the hitx and hity variables, create a new variable equal to the distance, in feet from home plate, of Cabrera’s base hits. (This variable is computed by simply applying the Pythagorean Theorem – remember that home plate is at the origin.)
```{r Ch6.Q4b}
pythag <- function(a, b){
  c2 <- a^2 + b^2
  sqrt(c2)
}
cabrera <- cabrera %>% mutate(distance = pythag(hitx, hity))
head(cabrera)
```
+ In the newly created data frame, create a gameDay variable indicating the day of the year (from 0 to 365) in which the game took place (see Section 6.2.6).
```{r Ch6.Q4c}
cabrera <- cabrera %>% 
  mutate(gameDay = as.integer(format(gamedate, format="%j")))
head(cabrera)
```
+ Build a scatterplot featuring gameDay on the x-axis, distance on the y-axis and a smooth line with error bands. Does the resulting plot appear to indicate changes in Cabrera’s power during the season?
```{r Ch6.Q4d}
ggplot(cabrera, aes(x = gameDay, y = distance)) +
  theme_bw() + 
  geom_point(size = 1.5) + 
  ylim(0, 550) + 
  xlim(90, 274) + 
  xlab("Distance") + 
  ylab("Day of year") + 
  stat_smooth(method = "lm", color = "red", size = 1.5) # Note that the ~4300 removed values are NAs.
# Looks like a slight decline in distance but I highly doubt it is statistically significant let alone means anything in reality.
```