---
title: "tidy Baseball Chapter 8"
author: "Chris Hamm"
date: "`r format(Sys.Date())`"
output:  
      html_document: 
        keep_md: TRUE
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

## Chapter 8 - *Career trajectories*


```{r preliminaries}
library("Lahman")
library("tidyverse"); options(dplyr.width = Inf)

set.seed(8761825)
devtools::session_info()
```

### Section 8.2 - *Mickey Mantel's Batting Trajectory*
```{r Sec_8.2}
# The Batting and Master files are already loaded with the Lahman package. 
head(Batting) # using read.csv here rather than read_csv because it imports triples as X3B rather than 3B, which is not the best
# I don't know how to recode variables with dplyr so I will use recode from "car."
Batting$SF <- car::recode(Batting$SF, "NA = 0")
Batting$HBP <- car::recode(Batting$HBP, "NA = 0")
head(Batting)
dim(Batting)

mantle.id <- Master %>% filter(nameFirst == "Mickey" & nameLast == "Mantle") %>% select(playerID)
mantle.id

get.birthyear <- function(player.id){
	playerline <- Master %>% 
	  filter(playerID == as.character(player.id))
	birthyear <- playerline$birthYear
	birthmonth <- playerline$birthMonth
	ifelse(birthmonth >= 7, birthyear + 1, birthyear)
}
get.birthyear(mantle.id)
```


#### Note that the book formula for OBP does not include HBP; if you exclude this you will get the incorrect result. The book inculdes hits in the denominator, which I have re-added as well.
```{r Sec_8.2b}
get.stats <- function(player.id){
	d <- Batting %>% filter(playerID == as.character(player.id))
	byear <- get.birthyear(as.character(player.id))
	d <- d %>% mutate(Age = yearID - byear, 
	             SLG = (((H - X2B - X3B - HR) + (2 * X2B) + (3 * X3B) + (4 * HR)) / AB), 
	             OBP = ((H + BB + HBP) / (AB + BB + SF + HBP)), 
	             OPS = SLG + OBP)
	return(d)
}
Mantle <- get.stats(mantle.id)
Mantle
```

### Figure 8.1
```{r Fig_8.1, fig.cap = "Scatterplot of OPS against age for Mickey Mantle."}
ggplot(Mantle, aes(x = Age, y = OPS)) + 
	theme_bw() + 
	geom_point(size = 3) +
	xlim(16, 39)
```

Create a smooth curve of the quadratic:
$A + B(Age − 30) + C(Age − 30)^2$ where $A$, $B$, and $C$ are constants:

  1. The constant $A$ is predicted by the value of OPS when the player reaches 30 years of age. 
  1. The function reaches its highest value at
    $PEAK.AGE = 30 - \frac{B}{2C}$. This value is estiamted to be the player's peak batting performance. 
  1. The maximum value of the curve is: $Max = A - \frac{B^2}{4C}$. This is the estimate of the largest OPS of the player over his career. 
  1. The $C$ coefficient explains the curve of the quadratic function and usually takes a negative value. "Large" values represent stronger curves (rapid rise and delcline). We'll use the "lm" function to fit the quadratic curve formula: $OPS \sim I(Age - 30) + I((Age - 30)^2)$

```{r Sec_8.2c}
fit.model <- function(d){
	fit <- lm(OPS ~ I(Age - 30) + I((Age - 30)^2), data = d)
	b <- coef(fit)
	Age.max <- 30 - b[2] / b[3] / 2
	Max <- b[1] - b[2]^2 / b[3] / 4
	list(fit = fit, Age.max = Age.max, Max = Max)
}

F2 <- fit.model(Mantle)
F2
summary(F2$fit)
```
This makes the model: `r cat(sprintf("%.02f %.02f(Age - 30) %.02f (Age - 30)^2", coef(F2$fit)[1], coef(F2$fit)[2], coef(F2$fit)[3]))`

```{r Fig_8.2, fig.cap = "Scatterplot of OPS against age for Mickey Mantle with a quadratic fit added. The location of the peak age and the maximum OPS fit are displayed."}
# I don't know how to add a function line in ggplot, so base R it is.
plot(x = Mantle$Age, y = Mantle$OPS, pch = 19, las = 1, , cex =1.5, ylab = "OPS", xlab = "Age", xlim = c(18, 37), ylim = c(0, 1.3))
lines(Mantle$Age, predict(F2$fit, Age = Mantle$Age), lwd = 3)
abline(v = F2$Age.max, lwd = 3, lty = 2, col = "grey")
abline(h = F2$Max, lwd = 3, lty = 2, col = "grey")
text(29.5, .72, "Peak.age" , cex = 2)
text(20, 1.175, "Max", cex = 2)
```

### In honor of A-Rod retiring this season.
```{r A-Rod}
Arod.id <- Master %>% filter(nameFirst == "Alex" & nameLast == "Rodriguez") %>% select(playerID)
get.birthyear(Arod.id)
Arod <- get.stats(Arod.id) #only through 2015
A2 <- fit.model(Arod)
A2 # Arod peaked at 28 for OPS but stayed OK
# ggplot(Arod, aes(x = Age, y = OPS)) + theme_bw() + geom_point(size = 2) 
plot(x = Arod$Age, y = Arod$OPS, las = 1, pch = 19, cex =1.5, ylab = "OPS", xlab = "Age", xlim = c(17, 40), ylim = c(0, 1.1))
lines(Arod$Age, predict(A2$fit, Age = Arod$Age), lwd = 3, lty = 2)
```

### Section 8.3 - *Comparing trajectories*
```{r Sec_8.3}
head(Fielding)

AB.totals <- Batting %>% group_by(playerID) %>% summarize(Career.AB = sum(AB, na.rm = TRUE))
head(AB.totals)

Batting <- full_join(Batting, AB.totals, by = "playerID")
head(Batting)
Batting.2000 <- Batting %>% filter(Career.AB >= 2000)
head(Batting.2000)
dim(Batting.2000)

find.position <- function(p){
	positions <- c("OF", "1B", "2B", "SS", "3B", "C", "P", "DH")
	d <- subset(Fielding, playerID == p)
	count.games <- function(po)
		sum(subset(d, POS == po)$G)
	FLD <- sapply(positions, count.games)
	positions[FLD == max(FLD)][1]
}


PLAYER <- as.character(unique(Batting.2000$playerID))
POSITIONS <- sapply(PLAYER, find.position)
length(POSITIONS)

Fielding.2000 <- data.frame(playerID = names(POSITIONS), POS = POSITIONS)
head(Fielding.2000)
Batting.2000 <- left_join(Batting.2000, Fielding.2000, by = "playerID") # Becuase there are duplicate column names they get suffixes. The only way I know how to remove these is clunky.

# Batting.2000 <- merge(Batting.2000, Fielding.2000)
head(Batting.2000)
dim(Batting.2000)
```

#### Calculating career statistics
```{r Sec_8.3.2}
C.totals <- Batting.2000 %>% group_by(playerID) %>% summarize(
	C.G = sum(G, na.rm = TRUE), 
	C.AB = sum(AB, na.rm = TRUE), 
	C.R = sum(R, na.rm = TRUE), 
	C.H = sum(H, na.rm = TRUE), 
	C.2B = sum(X2B, na.rm = TRUE), 
	C.3B = sum(X3B, na.rm = TRUE), 
	C.HR = sum(HR, na.rm = TRUE), 
	C.RBI = sum(RBI, na.rm = TRUE), 
	C.BB = sum(BB, na.rm = TRUE), 
	C.SO = sum(SO, na.rm = TRUE), 
	C.SB = sum(SB, na.rm = TRUE)) %>% 
  mutate(C.AVG = (C.H / C.AB), 	C.SLG = (((C.H - C.2B - C.3B - C.HR) + (2 * C.2B) + (3 * C.3B) + (4 * C.HR)) / C.AB))
head(C.totals)

C.totals <- inner_join(C.totals, Fielding.2000, by = "playerID")
head(C.totals) 

C.totals$Value.POS <- with(C.totals,
	ifelse(POS == "C", 240,
	ifelse(POS == "SS", 168,
	ifelse(POS == "2B", 132,
	ifelse(POS == "3B", 84,
	ifelse(POS == "OF", 48,
	ifelse(POS == "1B", 12, 0)))))))
head(C.totals)
```

#### Computing similarity scores
```{r Sec_8.3.3}
similar <- function(p, number){
	P <- subset(C.totals, playerID == p)
	C.totals$SS <- with(C.totals,
	1000 -
	floor(abs(C.G - P$C.G) / 20) -
	floor(abs(C.AB - P$C.AB) / 75) -
	floor(abs(C.R - P$C.R) / 10) -
	floor(abs(C.H - P$C.H) / 15) -
	floor(abs(C.2B - P$C.2B) / 5) -
	floor(abs(C.3B - P$C.3B) / 4) -
	floor(abs(C.HR - P$C.HR) / 2) -
	floor(abs(C.RBI - P$C.RBI) / 10) -
	floor(abs(C.BB - P$C.BB) / 25) -
	floor(abs(C.SO - P$C.SO) / 150) -
	floor(abs(C.SB - P$C.SB) / 20) -
	floor(abs(C.AVG - P$C.AVG) / 0.001) -
	floor(abs(C.SLG - P$C.SLG) / 0.002) -
	abs(Value.POS - P$Value.POS))
C.totals <- C.totals[order(C.totals$SS, decreasing = TRUE), ]
C.totals[1:number, ]
}
similar(as.character(mantle.id), 6) # This is nice and all but I bet that hierarchical cluster analysis could do a great job at this. 
```

#### Defining age, OBP, SLG, and OPS variables
```{r Sec_8.3.4}
collapse.stint <- function(d){
	d %>% group_by(playerID, yearID) %>% summarize(G = sum(G), AB = sum(AB), R = sum(R), H = sum(H), X2B = sum(X2B), X3B = sum(X3B), HR = sum(HR), RBI = sum(RBI), SB = sum(SB), CS = sum(CS), BB = sum(BB), SH = sum(SH), SF = sum(SF), HBP = sum(HBP), SLG = (((H - X2B - X3B - HR) + (2 * X2B) + (3 * X3B) + 4 * HR) / AB), OBP = ((H + BB + HBP) / (AB + BB + HBP + SF)), OPS = SLG + OBP, Career.AB = Career.AB[1], POS = POS[1]) # This is the correct OBP formula. Earlier it is incorrect.
} 
Batting.2000a <- Batting.2000 %>%
  group_by(playerID, yearID) %>%
  collapse.stint
head(Batting.2000a)
dim(Batting.2000a)

player.list <- as.character(unique(Batting.2000$playerID))
birthyears <- sapply(player.list, get.birthyear)
Batting.2000a <- merge(Batting.2000a, data.frame(playerID = player.list, Birthyear = birthyears))
Batting.2000a$Age <- with(Batting.2000a, yearID - Birthyear)
Batting.2000a <- Batting.2000a[complete.cases(Batting.2000a$Age), ]
head(Batting.2000a)
```

#### Fitting and plotting trajectories
```{r Sec_8.3.5}
fit.trajectory <- function(d){
  fit <- lm(OPS ~ I(Age - 30) + I((Age - 30)^2), data = d)
  data.frame(Age = d$Age, Fit = predict(fit, Age = d$Age))
}

plot.trajectories <- function(first, last, n.similar, ncol){
  
get.name <- function(playerid){
  d1 <- subset(Master, playerID == playerid)
  with(d1, paste(nameFirst, nameLast))
}

player.id <- subset(Master, nameFirst == first & nameLast == last)$playerID

player.id <- as.character(player.id)

player.list <- as.character(similar(player.id, n.similar)$playerID)

Batting.new <- subset(Batting.2000a, playerID %in% player.list)

F2 <- Batting.new %>%
  group_by(playerID) %>%
  fit.trajectory # note here the book calls for a function "plot.traj"
F2a <- merge(F2, data.frame(playerID = player.list, Name = sapply(as.character(player.list), get.name)))
print(ggplot(F2a, aes(x = Age, y = Fit)) +
  theme_bw() + 
  geom_line(size = 1.5) +
  facet_wrap(~ Name, ncol = ncol)) 
# return(Batting.new)
}

plot.trajectories("Mickey", "Mantle", n.similar = 6, ncol = 2)
plot.trajectories("Derek", "Jeter", n.similar = 9, ncol = 3)
```


