---
title: "tidy_Baseball_Ch2"
author: "C. A. Hamm"
date: "July 31, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Beginning with Chapter 2, because Chapter 1 doesn't have much for us. 


### Load packages, set seed, and note which package versions we will use.
```{r preliminaries}
library("Lahman")
library("dplyr"); options(dplyr.width = Inf)
library("ggplot2")
library("devtools")
library("readr")

set.seed(8761825)
session_info()
```

```{r Spahn}
Spahn <- read_csv("https://raw.githubusercontent.com/maxtoki/baseball_R/master/data/spahn.csv", col_names = TRUE) # Note the "raw." to get the unformatted .csv

# create the FIP stat
Spahn <- Spahn %>% mutate(FIP = ((13 * HR) + (3 * BB) -2 * SO) / IP)
head(Spahn)
pos <- order(Spahn$FIP)
head(Spahn[pos, c("Year", "Age", "W", "L", "ERA", "FIP")])

Spahn1 <- Spahn %>% filter(Tm == "BSN" | Tm == "MLN") %>% mutate(Tm = factor(Tm, levels = c("BSN", "MLN")))
head(Spahn1)
by(Spahn1[, c("W-L", "ERA", "WHIP", "FIP")], Spahn1$Tm, summary)
```


```{r Batting}
Batting <- read_csv("https://raw.githubusercontent.com/maxtoki/baseball_R/master/data/Batting.csv", col_names = TRUE)
head(Batting)
dim(Batting)

Batting.60 <- Batting %>% filter(yearID >= 1960 & yearID <= 1969)
head(Batting.60)
dim(Batting.60)
max(Batting.60$yearID)
min(Batting.60$yearID)

compute.hr <- function(pid){
	d <- Batting.60 %>% filter(playerID == pid)
	sum(d$HR)
}

# The book way with multiple steps and a function:
players <- unique(Batting.60$playerID)
system.time(S <- sapply(players, compute.hr))
head(S)
length(S) #1786 players had home runs in the 1960's

# The tidy way, one line of code:
system.time(S1 <- Batting.60 %>% group_by(playerID) %>% summarize(sum(HR)))
head(S1)
dim(S1)


# create a data frame 
dataframe.AB <- Batting %>% select(playerID, AB, HR, SO) %>% group_by(playerID) %>% filter(!is.na(AB)) %>% summarize(AB = sum(AB), HR = sum(HR), SO = sum(SO))

head(dataframe.AB)
dim(dataframe.AB)

# Note here that merging as described in the book makes no sense. Why add repeared rows of summarized data back to the original data frame "Batting"

Batting2 <- dplyr::full_join(Batting, dataframe.AB, by = "playerID")
head(Batting2)
dim(Batting2)


Batting.5000 <- dataframe.AB %>% filter(AB >= 5000)
dim(Batting.5000)
head(Batting.5000)


# This function "compute.hr" is now unnecessary, as we have already calculated the relevent statistics. See how, with one line of dplyr code, we can replace a function and other awkward subsetting.
Batting.5000 %>% filter(playerID == "aaronha01") %>% summarize(AB = sum(AB, na.rm = TRUE), HR = sum(HR, na.rm = TRUE), SO = sum(AB, na.rm = TRUE))

# Figure 2.8, where we want to plot the SO/AB (Y) against the HR/AB (X)
ggplot(Batting.5000, aes(x = HR / AB, y = SO / AB)) + theme_bw() + xlim(0, 0.1) + ylim(0, 0.35) + geom_point(cex = 2) + stat_smooth(method = "loess", col = "red") + ylab("SO / AB") + xlab("HR / AB")
```

### Chapter 2 exercises
1. Question 1
  + In R, place the stolen base, caught stealing, and game counts in the vectors SB, CS, and G.
  + For all players, compute the number of stolen base attempts SB + CS and store in the vector SB.Attempt.
  + For all players, compute the success rate Success.Rate = SB / SB.Attempt.
  + Compute the number of stolen bases per game SB.Game = SB / Game.
  + Construct a scatterplot of the stolen bases per game against the success rates. Are there particular players with unusually high or low stolen base success rates? Which player had the greatest number of stolen bases per game?

```{r Q2.1a}
# Import the Hall of Fame data set from the repo:
hof <- read_csv("https://raw.githubusercontent.com/maxtoki/baseball_R/master/data/hofbatting.csv", col_names = TRUE)
dim(hof)
head(hof)
hof$X2 <- gsub(" HOF", "", hof$X2)
head(hof$X2)

# a)
SB <- hof %>% select(SB) %>% arrange(desc(SB))
CS <- hof %>% select(CS) %>% arrange(desc(CS))
G <- hof %>% select(G) %>% arrange(desc(G))

# b)
SB.Attempt <- hof %>% select(SB, CS) %>% transmute(SB.Attempt = SB + CS) %>% arrange(desc(SB.Attempt))
head(SB.Attempt)

# c) 
Success.Rate <- hof %>% select(SB, CS) %>% transmute(Success.Rate = SB / (SB + CS)) %>% arrange(desc(Success.Rate))
head(Success.Rate)

# d)
SB.Game <- hof %>% select(SB, G) %>% transmute(SB.Game = SB / G) %>% arrange(desc(SB.Game))
head(SB.Game)

# e)
ggplot(hof, aes(x = (SB / (SB + CS)), y = (SB / G)), label = X2) + theme_bw() + geom_point(size = 1.5) + xlab("Stolen Base \nSuccess Rate") + ylab("Stolen Bases \nper Game") + geom_text(size= 2, aes(label = X2), nudge_y = 0.0125)
``` 
