---
title: "tidy Baseball Chapter 4"
author: "C. A. Hamm"
date: "`r format(Sys.Date())`"
output:  
      html_document:  
        keep_md: true  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Chapter 4 -  The relation between runs and wins

```{r preliminaries}
library("Lahman")
library("dplyr"); options(dplyr.width = Inf)
library("ggplot2")
library("devtools")
library("readr")

set.seed(8761825)
session_info()
```


```{r Teams_Table}
setwd("~/Desktop/Projects/Baseball_R")

teams <- read_csv("2016_core/core/Teams.csv", col_names = TRUE)

tail(teams)
dim(teams)


## Section 4.2 - Teams in Lahman's Database
# run differential is the difference between runs scored and runs allowed. Rather than use the with statements in section 4.2 we will use a dplyr pipeline

myteams <- teams %>% filter(yearID > 2000) %>% 
  select(teamID, yearID, lgID, G, W, L, R, RA) %>% 
  mutate(RD = R - RA, Wpct = W / (W + L))
tail(myteams)

# Figure 4.1
ggplot(myteams, aes(x = RD, y = Wpct)) + 
  geom_point(cex = 2, alpha = 0.6) + 
  theme_bw() + 
  stat_smooth(method = lm, colour = "black") + 
  xlab("Run differential") + 
  ylab("Winning percentage")
```


```{r linear_regression}
linfit <- lm(Wpct ~ RD, data = myteams)
summary(linfit)
# Wpct = 0.50 + 0.000628*RD

myteams <- myteams %>% mutate(linWpct = predict(linfit), linResiduals = residuals(linfit))
tail(myteams)
min(myteams$linResiduals)
max(myteams$linResiduals)

# Figure 4.2
ggplot(myteams, aes(x = RD, y = linResiduals)) + 
  geom_point(size = 2) + 
  theme_bw() + 
  ylab("Residuals") + 
  xlab("Run differential") + 
  scale_y_continuous(limits = c(-0.09, 0.09)) + 
  scale_x_continuous(limits = c(-400, 400)) 

mean(myteams$linResiduals)
linRMSE <- sqrt(mean(myteams$linResiduals^2))
linRMSE

nrow(subset(myteams, abs(linResiduals) < linRMSE)) / nrow(myteams)
nrow(subset(myteams, abs(linResiduals) < 2 * linRMSE)) / nrow(myteams)

# section 4.4 - the Pythagorean expectation
myteams <- myteams %>% 
  mutate(pytWpct = R^2 / (R^2 + RA^2), 
         pytResiduals = Wpct - pytWpct)
tail(myteams)
sqrt(mean(myteams$pytResiduals^2))

ggplot(myteams, aes(y = pytWpct, x = Wpct)) +
  theme_bw() + 
  geom_point(cex = 2, alpha = 0.6) + 
  ylab("Pythagorean Win %") + 
  xlab("Win %") + 
  stat_smooth(method = "lm", color = "black")

lm2 <- lm(pytWpct ~ Wpct, data = myteams)
summary(lm2) # 0.034 + 0.93*Wpct

# a function to compare the Bill James Pythagorean model with the linear model. RunsSc = Runs Scored, RunsAll = Runs Allowed, N = exponant
Pythag.line <- function(RunsSc, RunsAll, N){
	linear <- 0.50 + 0.000628 * (RunsSc - RunsAll)
	Pythag <- RunsSc^N / (RunsSc^N + RunsAll^N)
	return(list = c("linear" = linear, "Pythagorean" = Pythag))
}
Pythag.line(1620, 810, N = 2)
Pythag.line(186, 0, N = 2)
Pythag.line(100, 100, N = 2)

# Section 4.5 - the exponent of the Pythagorean formula
myteams <- myteams %>% 
  mutate(logWratio = log(W / L), 
         logRratio = log(R / RA))
tail(myteams)

pytFit <- lm(logWratio ~ logRratio, data = myteams)
summary(pytFit) # suggests a Pythagorean exponent of 1.88 rather than 2.

# Section 4.6 - Good and Bad predictions
games1 <- Pythag.line(875, 737, N = 2)
games1 * 162
games1a <- Pythag.line(875, 757, N = 1.88)
games1a * 162


# Section 4.6 - Good and bad predictions based on the Pythagorean Formula
gl2011 <- read_delim("baseball_R-master/data/gl2011.txt", delim = ",", col_names = FALSE)
head(gl2011) # needs headers
glheaders <- read_csv("baseball_R-master/data/game_log_header.csv")
glheaders
names(gl2011) <- names(glheaders)
head(gl2011)

BOS2011 <- gl2011 %>% filter(HomeTeam == "BOS" | VisitingTeam == "BOS") %>% 
  select(VisitingTeam, HomeTeam, VisitorRunsScored, HomeRunsScore) %>% 
  mutate(ScoreDiff = ifelse(HomeTeam == "BOS", yes = HomeRunsScore - VisitorRunsScored, no = VisitorRunsScored - HomeRunsScore), 
         W = ScoreDiff > 0)
head(BOS2011)
# aggregate(abs(BOS2011$ScoreDiff), list(W = BOS2011$W), summary)

results <- gl2011 %>% 
  select(VisitingTeam, HomeTeam, VisitorRunsScored, HomeRunsScore) %>% 
  mutate(winner = ifelse(HomeRunsScore > VisitorRunsScored, yes = as.character(HomeTeam), no = as.character(VisitingTeam)), 
         diff = abs(VisitorRunsScored - HomeRunsScore))
head(results)

onerungames <- results %>% filter(diff == 1)
dim(onerungames)
head(onerungames)

onerunwins <- onerungames %>% 
  group_by(winner) %>% 
  tally()
names(onerunwins) <- c("teamID", "onerunW")
onerunwins

teams2011 <- myteams %>% filter(yearID == 2011)
teams2011[teams2011$teamID == "LAA", "teamID"] <- "ANA"
teams2011 <- merge(teams2011, onerunwins)
head(teams2011)

# Figure 4.3
ggplot(teams2011, aes(x = onerunW, y = pytResiduals)) + geom_point(cex = 2) + theme_bw() + xlab("One run wins") + ylab("Pythagorean residuals")

pit <- read_csv("baseball_R-master/data/pitching.csv", col_names = TRUE)
head(pit)

top_closers <- pit %>% filter(GF > 50, ERA < 2.5) %>% select(playerID, yearID, teamID)
head(top_closers)
teams_top_closers <- merge(myteams, top_closers)
summary(teams_top_closers$pytResiduals)
mean(teams_top_closers$pytResiduals) * 162 # teams with a top closer will outperform their Pythagorean expectation by 0.8 games


# Section 4.7 -  How many runs for a win?
IR <- function(RS, RA){IRtable <- expand
	round((RS^2 + RA^2)^2 / (2 * RS * RA^2), 1)
}
IRtable <- expand.grid(RS = seq(3, 6, 0.5), RA = seq(3, 6, 0.5))
dim(IRtable)
rbind(head(IRtable), tail(IRtable))
IRtable <- IRtable %>% mutate(IRW = IR(RS, RA))
head(IRtable)

xtabs(IRW ~ RS + RA, data = IRtable)


